Inducing Domain Specific Languages for Bayesian Program Learning

### 2018-02-03 paper.tex
text primitives
started writing circuit section
vectorize mlp; changes to intro; small edits

Our algorithm is best understood as inference in a hierarchical Bayesian model (Figure~\ref{graphicalModel}: solid lines):
A DSL $\mathcal{D}$
samples a latent program $p$
which determines the agent's observations, $x$.
This makes our work in instance of
 \emph{Bayesian Program
  Learning} (BPL; see~\citep{lake2013one,Dechter:2013:BLV:2540128.2540316,ellis2016sampling,DBLP:conf/icml/LiangJK10}).
In tandem we learn a learn a neural network $q$ that regresses
from observations to programs.
We will show that the interaction of
the generative and recognition models is important for solving these program induction problems.

TODO: talk explicitly about what ideas we are borrowing from EC
%% We cast these problems as \emph{Bayesian Program
%%   Learning} (BPL; see~\citep{lake2013one,ellis2016sampling,DBLP:conf/icml/LiangJK10}),
%% where the goal is to infer from an observation $x$ a posterior distribution over programs, $\probability[p|x]$.
%% A DSL $\mathcal{D}$ specifies the vocabulary in which programs $p$ are written.
%% We equip our DSLs with a \emph{weight vector} $\theta$; together, $(\mathcal{D},\theta)$
%% define a probabilistic generative model over programs, $\probability[p|\mathcal{D},\theta]$.
%% In this BPL setting, $\probability[p|x]\propto \probability[x|p]\probability[p|\mathcal{D},\theta]$,
%% where the likelihood $\probability[x|p]$ is domain-dependent.
%% The solid lines in Fig.~\ref{graphicalModel} the diagram this generative model.
%% Alongside this generative model,
%% we infer a bottom-up recognition model, $q(x)$, which is a neural network that regresses from observations to a distribution over programs.

\begin{figure}
  \begin{tikzpicture}

  \node[latent] at (1,3) (d){$\mathcal{D}$};
  \node[latent] at (2.5,3) (t){$\theta$};
  \node[latent] at (1,1.5) (z){$p$};
  \node[latent] at (3.5,1.5) (tx){$\theta^{(x)}$};
  \node[obs] at (1,-0.5) (x) {$x$};
  \edge {z}{x};
  \edge {d,t}{z};
  \plate {}{(tx)(z)(x)}{$x\in X$};
  \draw [->,dashed] (x.east) to[out = 0,in = -90] node(nn)[fill = white]{
          \includegraphics[width = 1.5cm]{figures/MLP.png} 
  } (tx.south);
  \node at (nn.south) {$q(x)$};
  \draw [->,dashed] (d.west) to[out = 180,in = 180] (z.west);
  \draw [->,dashed] (tx.west) -- (z.east);

@@

@@

We frame this problem as maximum a posteriori (MAP) inference in the generative model diagrammed in Fig.~\ref{graphicalModel}. We wish to maximize the MAP probability of $\mathcal{D}$:
\begin{equation*}
  \probability[\mathcal{D}|X]\propto \probability[\mathcal{D}] \int \mathrm{d}\theta P(\theta|\mathcal{D})\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]
\end{equation*}
In general this marginalization over $\theta$ is intractable, so we make an AIC-style approximation\footnote{Sec.~\ref{grammarInductionSection} explains that $\mathcal{D}$ is a context-sensitive grammar.
Conventional NLP approaches to using variational inference to lower bound the marginal over $\theta$ do not apply in our setting.}, $A\approx \log\probability[\mathcal{D}|X] $:
\begin{align}
  A =   \log \probability[\mathcal{D}] + \argmax_{\theta}& \sum_{x\in X}\log \sum_p\probability[x|p]\probability[p|\mathcal{D},\theta]\nonumber\\
&+  \log P(\theta|\mathcal{D}) - ||\theta||_0 \label{AIC}
  \end{align}

@@

We maximize $J$ by alternatingly maximizing it w.r.t. the DSL and the frontiers:
%So interleaved with the DSL induction steps are program induction steps. Section~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{Program Search: Maxing $J$ w.r.t. the frontiers.} Here we
want to find new programs to add to  the frontiers so that $J$ increases the most.
Adding new programs to the frontiers means searching for new programs $p$ for task $x$
where $\probability[x,p|\mathcal{D},\theta]$ is large.
\\\noindent \textbf{DSL Induction: Maxing $J$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed and so we can evaluate $J$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $J$.

@@

\subsection{Boolean circuits}
As a toy warm-up domain, we consider the problem of learning
to build the Boolean circuits out of logic gates.
Now, if we equipped our agent with a full repertoire of
circuit primitives -- XOR gates, multiplexers, etc. -- this problem would be easy,
and we could use sophisticated symbolic search algorithms like SMT solvers to synthesize circuits.
Instead, we gave \system only a NAND gate,
and tasked it with building a set of 500 random circuits made out of
AND, NOT, and OR gates.
This experimental setup was introduced in the original EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316}.

Figure~\ref{circuitLearningCurve}

@@

@@

\begin{figure}
      \includegraphics[width = 9cm]{figures/polynomial.png}
\end{figure}
\subsection{String editing}\label{textSection}
\begin{figure}

>>>>

\begin{figure}
  \centering

  \begin{tikzpicture}

  \node[latent] at (1,3) (d){$\mathcal{D}$};
  \node[latent] at (2.5,3) (t){$\theta$};
  \node[latent] at (1,1.5) (z){$p$};
  \node[latent] at (3.5,1.5) (tx){$\theta^{(x)}$};
  \node[obs] at (1,-0.5) (x) {$x$};
  \edge {z}{x};
  \edge {d,t}{z};
  \plate {}{(tx)(z)(x)}{$x\in X$};
  \draw [->,dashed] (x.east) to[out = 0,in = -90] node(nn){} (tx.south);
  \draw [->,dashed] (tx.west) -- (z.east);

  \node at (nn) {
    \begin{tikzpicture}[x=2.5cm,y=1.25cm,transform canvas={scale=0.2,shift={+(-1,2.5)}}]
      \tikzstyle{neuron}=[circle,fill=blue!50,minimum size=20pt]
      \fill[fill=white] (-0.25,-0.5) rectangle (2.25,-4.5);
      \node[rectangle,width=2cm] at (1,1) {};
      \foreach \name / \y in {1,...,4}
          \node[neuron] (I-\name) at (0,-\y) {};
      \foreach \name / \y in {1,...,3}
          \node[neuron] (H-\name) at (1,-\y-0.5) {};
      \foreach \name / \y in {1,...,4}
          \node[neuron] (O-\name) at (2,-\y) {};
      \foreach \source in {1,...,4}
          \foreach \dest in {1,...,3}
              \draw [-latex] (I-\source) -- (H-\dest);
      \foreach \source in {1,...,3}
          \foreach \dest in {1,...,4}
              \draw [-latex] (H-\source) -- (O-\dest);
    \end{tikzpicture}
  };
  \node[shift={+(0,-0.55)}] at (nn) { $q(x)$ };

@@

Our algorithm is best understood as inference in a hierarchical Bayesian model (Figure~\ref{graphicalModel}: solid lines):
A DSL $\mathcal{D}$
determines a distribution with production probabilities $\theta$,
from which a latent program $p$ is sampled,
which finally determines the agent's observations, $x$.
This makes our work in instance of
 \emph{Bayesian Program
  Learning} (BPL; see~\citep{lake2013one,Dechter:2013:BLV:2540128.2540316,ellis2016sampling,DBLP:conf/icml/LiangJK10}).
In tandem we learn a neural network $q$ that regresses from observations to
programs by proposing a distribution over programs $\theta^{(x)} = q(x)$.
We will show that the interaction of the generative and recognition models
is important for solving these program induction problems.

TODO: talk about existing works as special cases of, and points of inspiration for, \system:
\begin{itemize}
  \item Maximum-likelihood estimation of a PCFG \citep{DBLP:conf/icml/LiangJK10} with
    production probabilities $\theta$, a fixed the DSL $\mathcal{D}$ and no neural network
    $q$;
    % although some of the Liang paper actually learns subexpressions and therefore DSL
  \item The EC algorithm \citep{Dechter:2013:BLV:2540128.2540316} has no recognition model
    $q$;
    % because this is a primary inspiration, might be worth further mention:
    % different program representation (routing vs. substitution; simple vs polymorphic types);
    % different grammar induction (sequitur vs. fragment grammar)
  \item DeepCoder and RobustFill \citep{balog2016deepcoder,devlin2017robustfill} both fix the
    DSL $\mathcal{D}$ and does not use $\theta$ in favor of solely relying on the neural
    network $q$.
    % although RobustFill has very different enumeration
\end{itemize}

%% We cast these problems as \emph{Bayesian Program
%%   Learning} (BPL; see~\citep{lake2013one,ellis2016sampling,DBLP:conf/icml/LiangJK10}),
%% where the goal is to infer from an observation $x$ a posterior distribution over programs, $\probability[p|x]$.
%% A DSL $\mathcal{D}$ specifies the vocabulary in which programs $p$ are written.
%% We equip our DSLs with a \emph{weight vector} $\theta$; together, $(\mathcal{D},\theta)$
%% define a probabilistic generative model over programs, $\probability[p|\mathcal{D},\theta]$.
%% In this BPL setting, $\probability[p|x]\propto \probability[x|p]\probability[p|\mathcal{D},\theta]$,
%% where the likelihood $\probability[x|p]$ is domain-dependent.
%% The solid lines in Fig.~\ref{graphicalModel} the diagram this generative model.
%% Alongside this generative model,
%% we infer a bottom-up recognition model, $q(x)$, which is a neural network that regresses from observations to a distribution over programs.

@@

We frame this problem as maximum a posteriori (MAP) inference in the generative model diagrammed in Fig.~\ref{graphicalModel}. We wish to maximize the MAP probability of $\mathcal{D}$:
\begin{equation*}
  \probability[\mathcal{D}|X]\propto \probability[\mathcal{D}] \int \mathrm{d}\theta P(\theta|\mathcal{D})\prod_{x\in X} \sum_p \probability[x|p]\probability[p|\mathcal{D},\theta]
\end{equation*}
In general this marginalization over $\theta$ is intractable, so we make an AIC-style approximation\footnote{Sec.~\ref{grammarInductionSection} explains that $\mathcal{D}$ is a context-sensitive grammar.
Conventional natural-language processing (NLP) approaches to using variational inference to lower bound the marginal over $\theta$ do not apply in our setting.}, $A\approx \log\probability[\mathcal{D}|X] $:
\begin{align}
  A =   \log \probability[\mathcal{D}] + \argmax_{\theta}& \sum_{x\in X}\log \sum_p\probability[x|p]\probability[p|\mathcal{D},\theta]\nonumber\\
&+  \log P(\theta|\mathcal{D}) - ||\theta||_0 \label{AIC}
  \end{align}

@@

We maximize $J$ by alternating maximization w.r.t. the DSL and the frontiers:
%So interleaved with the DSL induction steps are program induction steps. Section~\ref{explorationSection}
%explains how \system synthesizes new programs and given a DSL.
\\\noindent \textbf{Program Search: Maxing $J$ w.r.t. the frontiers.} Here we
want to find new programs to add to  the frontiers so that $J$ increases the most.
Adding new programs to the frontiers means searching for new programs $p$ for task $x$
where $\probability[x,p|\mathcal{D},\theta]$ is large.
\\\noindent \textbf{DSL Induction: Maxing $J$ w.r.t. the DSL.} Here $\left\{\mathcal{F}_x \right\}_{x\in X}$ is held fixed and so we can evaluate $J$. Now the problem is that of searching the discrete space of DSLs and finding one maximizing $J$.

@@

\subsection{Boolean circuits}
As a toy warm-up domain, we consider the problem of learning
to build the Boolean circuits out of logic gates.
Now, if we equipped our agent with a full repertoire of
circuit primitives -- XOR gates, multiplexers, etc. -- this problem would be easy,
and we could use sophisticated symbolic search algorithms like SMT solvers to synthesize circuits.
Instead, we gave \system only a NAND gate,
and tasked it with building a set of 500 random circuits made out of
AND, NOT, and OR gates.
This experimental setup was introduced in the original EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316}.
For this domain our recognition model is a simple multilayer perceptron
whose inputs are the input/output examples.
Figure~\ref{circuitLearningCurve} shows our algorithm learning to build circuits; Table~\ref{circuitExperimentTable}

@@

\begin{table}
\begin{tabular}{ll}
  \toprule
  Primitives&\texttt{nand}

\\\midrule
  Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  \midrule
  Likelihood $\probability[x|p]$& $\prod_n \indicator[p(i_n) = o_n]$
  \\\midrule
  \begin{tabular}{l}
    Subset of \\Learned DSL
    \end{tabular}&

  \\\bottomrule
\end{tabular}
\caption{}\label{circuitExperimentTable}
\end{table}

@@

\begin{figure}
      \includegraphics[width = 9cm]{figures/polynomial.png}
\end{figure}
\subsection{String editing}\label{textSection}

\begin{tabular}{ll}
  \toprule
  Primitives&\begin{tabular}{l}
    \texttt{0}, \texttt{+1},\texttt{-1}\\
    \texttt{split}, \texttt{join}, \texttt{map}, \texttt{length}, \texttt{slice}\\
    \texttt{append}, \texttt{chr->str}\\
    \texttt{lowercase}, \texttt{uppercase}, \texttt{capitalize}\\
    \texttt{""},    \texttt{","},     \texttt{" "},    \texttt{"<"}, \texttt{">"}
    \end{tabular}

\\\midrule
  Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  \midrule
  Likelihood $\probability[x|p]$& $\prod_n \indicator[p(i_n) = o_n]$
  \\\midrule
  \begin{tabular}{l}
    Subset of \\Learned DSL
    \end{tabular}&

  \\\bottomrule
  \end{tabular}
\begin{figure}
